---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Setting Global Options
```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

## Loading and preprocessing the data
Importing tidyverse, dataset and converting date.
```{r importing}
library(tidyverse)
df <- read.csv("activity.csv")
df$date <- as.Date(df$date)
```

## What is mean total number of steps taken per day?
Aggregating data to daily frequency and printing summary.
```{r mean_summary}
df_daily <- df %>% 
    filter(!is.na(steps)) %>% 
    group_by(date) %>% 
    summarise(total_steps = sum(steps, na.rm = TRUE))

df_daily_summary <- df_daily %>% 
    summarise(mean_total = mean(total_steps, na.rm = TRUE),
              median_total = median(total_steps, na.rm = TRUE))

print(paste("mean total number of steps taken per day:", round(df_daily_summary[[1]], 0)))
print(paste("median total number of steps taken per day:", round(df_daily_summary[[2]], 0)))
```

Histogram of Daily Steps
```{r plot1}
g <- ggplot(df_daily, aes(x = total_steps)) +
    geom_histogram(bins = 20, fill = "skyblue", color = "black") +
    labs(title = "Histogram of Daily Steps",
         x = "Total Daily Steps",
         y = "Count")
print(g)
```

## What is the average daily activity pattern?
Average Daily Activity Pattern time series plot.
```{r plot2}
df_interval <- df %>% 
    group_by(interval) %>% 
    summarise(avg_steps = mean(steps, na.rm = TRUE)) %>% 
    arrange(interval)

g2 <- ggplot(df_interval, aes(x = interval, y = avg_steps)) +
    geom_line() +
    labs(
        title = "Average Daily Activity Pattern",
        x = "5-minute Interval",
        y = "Average Steps")

print(g2)
```

5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.

```{r interval_max}
interval_max <- df_interval %>% 
    slice_max(avg_steps, n = 1) %>% 
    pull(interval)
print(paste("interval which has max average steps:", interval_max))
```

## Imputing missing values

#### Strategy for imputing NAs
- I'm going to impute NAs by using the 5-min interval averages derived above. The 5-min interval will help retain the shape of the activities throughout the day.
- The process will use the 5-min interval averages if NA is present, otherwise the original steps will be used in a newly derived column called steps_filled.

```{r imputation}
missing_values <- df %>% 
    filter(is.na(steps)) %>% 
    nrow()

print(paste("number of missing values:", missing_values))

df_imputed <- df %>% 
    left_join(df_interval, by = "interval") %>% 
    mutate(steps_filled = if_else(is.na(steps), avg_steps, steps))

df_daily_imputed <- df_imputed %>% 
    group_by(date) %>% 
    summarise(total_steps = sum(steps_filled, na.rm = TRUE))
```

Histogram after imputation.
Compared to the original histogram without imputation of NAs, the histogram below shows more counts in the mean/median step range. This is as expected as the NAs were imputed using the 5-min interval averages.

```{r plot3}
g3 <- ggplot(df_daily_imputed, aes(x = total_steps)) +
    geom_histogram(bins = 20, fill = "skyblue", color = "black") +
    labs(title = "Histogram of Daily Imputed Steps",
         x = "Total Daily Imputed Steps",
         y = "Count")
print(g3)
```

Summary of mean and median after imputation.
```{r imputed_summary}
df_daily_imputed_summary <- df_daily_imputed %>% 
    summarise(mean_total = mean(total_steps, na.rm = TRUE),
              median_total = median(total_steps, na.rm = TRUE))

print(paste("Summary of mean imputation:", round(df_daily_imputed_summary[[1]], 0)))

print(paste("Summary of median imputation:", round(df_daily_imputed_summary[[2]], 0)))
```

## Are there differences in activity patterns between weekdays and weekends?

Yes, there are noticeable differences in activity patterns:

- **Weekends show higher overall activity during the daytime**, particularly between **10 AM and 5 PM**.
- **Weekdays have a higher peak 5-minute interval**, with a maximum of about **230 steps at interval 835**, compared to about **167 steps at interval 915** on weekends.

```{r}
df_imputed <- df_imputed %>% 
    mutate(
        day_type = if_else(
            wday(date, label = TRUE) %in% c("Sat", "Sun"),
            "weekend",
            "weekday"),
        day_type = factor(day_type, levels = c("weekend", "weekday")))
```

```{r plot4}
df_imputed_day_type <- df_imputed %>% 
    group_by(interval, day_type) %>% 
    summarise(avg_steps = mean(steps_filled, na.rm = TRUE)) %>% 
    arrange(interval, day_type)

g4 <- ggplot(df_imputed_day_type, aes(x = interval, y = avg_steps)) +
    geom_line() +
    facet_wrap(~ day_type, ncol = 1) +
    labs(
        title = "Average Daily Activity Pattern by Day Type",
        x = "5-minute Interval",
        y = "Average Steps"
    )

print(g4)
```